---
title: "Childcare Cost Burdens in Georgia and Washington, DC"
author: "Jennifer Watkins"
format: html
execute:
  echo: false
  warning: false
  message: false
  output: true
---
```{r}
# packages
library(tidyverse)
library(janitor)
library(scales)

#data
childcare <- read_csv("childcare_infants.csv")

#data frames
childcare_cost_burden <- childcare |> 
  mutate(median_center_annual = (median_center_infant*52),
         median_family_annual = (median_family_infant*52),
         percent_income_family = (median_family_annual/median_income_2018)*100,
         percent_income_center = (median_center_annual/median_income_2018)*100)

DC_childcare_cost_burden <- childcare_cost_burden |> 
  filter(state_abbreviation=="DC") |> 
  drop_na(percent_income_family,percent_income_family)

GA_childcare_cost_burden <- childcare_cost_burden |>
  filter(state_abbreviation=="GA") |> 
  drop_na(percent_income_family,percent_income_family) |> 
  summarize(percent_income_family_ga=median(percent_income_family),
            percent_income_center_ga=median(percent_income_center),
            median_income_2018_ga=median(median_income_2018),
            state_name=unique(state_name)) |> 
  rename(percent_income_family=percent_income_family_ga,
         percent_income_center=percent_income_center_ga,
         median_income_2018=median_income_2018_ga)

cost_burden_by_state <- bind_rows(DC_childcare_cost_burden,GA_childcare_cost_burden) |> 
  select(state_name,median_income_2018,percent_income_family,percent_income_center) |> 
  pivot_longer(cols=c(percent_income_family,percent_income_center),
               names_to="care_type",
               names_prefix="percent_income_",
               values_to="percentage_of_income") |> 
  mutate(percentage_of_income=percentage_of_income/100)

percentage_of_income_formatted <- cost_burden_by_state |> 
  mutate(percentage_of_income=percent(percentage_of_income,accuracy=.1))
```
# Definitions
Median income
: Median household income expressed in 2018 dollars. Used as the basis for percentage of income spent on child care.

Median center cost
: Median price charged for Center-based Care for infants (i.e. aged 0 through 23 months). A child care center is a facility outside a private home and is more likely to split childred up by age groups. Centers usually have a larger total group of children than family care.

Median family cost
: Full-time median price charged for Family Childcare for infants (i.e. aged 0 through 23 months). Family childcare is provided in a person's private home and are more likely to have mixed ages in one group of children. Family childcare usually has a smaller overall group of children.

# Results

* The percentage of a family's income spent on childcare is tied to the average income of the state, even across providers.
  + For example, Georgia has a lower median income than DC, and has a significantly lower percentage of income spent on childcare
* Families spent more on center-based childcare across states than they do when childcare is provided by family members.

#### Fig. 1a
```{r}
# data viz

ggplot(data=percentage_of_income_formatted,
       mapping=aes(x=care_type,
                   y=percentage_of_income,
                   fill=state_name,
                   label=percentage_of_income))+
  geom_col(position="dodge")+
  geom_label(vjust=1)+
  labs(title="Income spent on childcare by state and provider",
       caption="GA annual income = $40,768
       DC annual income= $70,693",
       x="Type of Childcare Provider",
       y="Percentage of Income Spent",
       fill="State")+
  theme_minimal()
```

